---
title: "Cholera data exploration"
author: "Seth Caldwell"
date: "2023-05-15"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE)

library(tidyverse)
library(gghdx)
gghdx()

output_dir <- file.path(
  Sys.getenv("CERF_GM_DIR"),
  "cholera_exploration",
  "outputs"
)

df_cerf <- read_csv(
  file.path(
    output_dir,
    "cerf_cholera_allocations.csv"
  )
)

df_cholera <- read_csv(
  file.path(
    output_dir,
    "cholera_cases.csv"
  )
)

```

## Cholera exploration

CERF has been scraping data from WHO AFRO emergency bulletins. We want to see
how this data compares with CERF allocations, as we don't currently have an
additional source of cholera case data.

```{r, fig.height = 8}
ggplot() +
  geom_line(
    data = df_cholera,
    mapping = aes(
      x = date,
      y = cholera_cases
    )
  ) +
  geom_segment(
    data = df_cerf,
    mapping = aes(
      x = date,
      xend = date,
    ),
    y = 0,
    yend = Inf
  ) +
  facet_wrap(
    ~ country,
    ncol = 3,
    scales = "free_y"
  ) +
  scale_y_continuous(
    labels = scales::label_comma()
  ) +
  scale_x_date(
    date_breaks = "2 years",
    date_labels = "%Y"
  ) +
  labs(
    x = "",
    y = "Cholera cases",
    title = "Cholera cases and CERF cholera allocations",
    subtitle = "CERF allocations as vertical lines"
  ) +
  theme(
    strip.text.x = element_text(size = 10),
  )
```

From the above, we can definitely see a relationship between cholera cases and CERF allocations (obviously) and it isn't always the case that CERF is allocating after to the peak of an outbreak. What if we activated if there was a first case reported after no cases and once the total caseload reaches 1,000, 5,000 or 10,000 cases.

```{r, fig.height = 8}
# alert helper function
# only generate alert when boundary crossed from below
lim_alert <- function(x, lim) {
  x >= lim & lag(x) < lim
}

# create alerts dataset
df_alert <- df_cholera %>%
  group_by(
    iso3
  ) %>%
  mutate(
    alert_first = cholera_cases > 0 & lag(cholera_cases) == 0,
    alert_1k = lim_alert(cholera_cases, 1000),
    alert_5k = lim_alert(cholera_cases, 5000),
    alert_10k = lim_alert(cholera_cases, 10000)
  ) %>%
  rowwise() %>%
  mutate(
    alert_any = any(c_across(starts_with("alert")))
  ) %>%
  ungroup()

# plot the results
ggplot() +
  geom_line(
    data = df_cholera,
    mapping = aes(
      x = date,
      y = cholera_cases
    )
  ) +
  geom_segment(
    data = df_alert %>% filter(alert_any),
    mapping = aes(
      x = date,
      xend = date,
    ),
    y = 0,
    yend = Inf
  ) +
  facet_wrap(
    ~ country,
    ncol = 3,
    scales = "free_y"
  ) +
  scale_y_continuous(
    labels = scales::label_comma()
  ) +
  scale_x_date(
    date_breaks = "2 years",
    date_labels = "%Y"
  ) +
  labs(
    x = "",
    y = "Cholera cases",
    title = "Cholera cases and potential alerts",
    subtitle = "Alerts on vertical lines"
  ) +
  theme(
    strip.text.x = element_text(size = 10),
  )
```
Here, the alerts are potentially too frequent in some cases, and not always due to rapid rises in caseload, such as in Ethiopia and Mozambique. Let's limit the alerts to only happening if that particular alert also hasn't activated in the past month, to reduce this frequency, but still allow rapid alerts if the rise in caseloads is particularly rapid or alarming.

```{r, fig.height = 8}
df_alert_reduced <- df_alert %>%
  mutate(
    across(
      .cols = alert_first:alert_10k,
      .fns = ~ zoo::rollsumr(x = .x, k = 30, fill = 0) == 1 & .x
    )
  ) %>%
  rowwise() %>%
  mutate(
    alert_any = any(c_across(alert_first:alert_10k))
  ) %>%
  ungroup()

# plot the results
p_alerts <- ggplot() +
  geom_line(
    data = df_cholera,
    mapping = aes(
      x = date,
      y = cholera_cases
    )
  ) +
  geom_segment(
    data = df_alert_reduced %>% filter(alert_any),
    mapping = aes(
      x = date,
      xend = date,
    ),
    y = 0,
    yend = Inf
  ) +
  facet_wrap(
    ~ country,
    ncol = 3,
    scales = "free_y"
  ) +
  scale_y_continuous(
    labels = scales::label_comma()
  ) +
  scale_x_date(
    date_breaks = "2 years",
    date_labels = "%Y"
  ) +
  labs(
    x = "",
    y = "Cholera cases",
    title = "Cholera cases and potential alerts",
    subtitle = "Alerts on vertical lines (only if not alerted in past 30 days)"
  ) +
  theme(
    strip.text.x = element_text(size = 10),
  )
```

So, it seems like we would not be activating particularly frequently. And the frequency
of alerts as designed relates to faster increases in cases which often grow much
further. But let's test this, and also see how these alerts relate to CERF allocations
when they were made.

First, just look at the simple relationship.

```{r, fig.height = 8}
p_alerts +
  geom_segment(
    data = df_cerf,
    mapping = aes(
      x = date,
      xend = date,
    ),
    y = 0,
    yend = Inf,
    color = hdx_hex("mint-hdx"),
    linewidth = 1
  ) +
  labs(
    title = "Cholera cases, potential alerts, and CERF allocations",
    subtitle = "Alerts in black, CERF allocations in green"
  )
```

```{r}
# get alerts in long format
df_alerts_long <- df_alert_reduced %>%
  filter(
    alert_any
  ) %>%
  select(
    -country
  ) %>%
  pivot_longer(
    alert_first:alert_10k,
    names_prefix = "alert_"
  ) %>%
  filter(
    value
  )

# connect with the CERF data
df_days_cerf <- left_join(
  rename(df_cerf, cerf_date = date),
  df_alerts_long,
  by = "iso3",
  relationship = "many-to-many"
) %>%
  filter(
    abs(cerf_date - date) <= 180
  ) %>%
  mutate(
    days_from_allocation = date - cerf_date
  ) %>%
  group_by(cerf_date) %>%
  mutate(
    earliest_day = min(days_from_allocation)
  ) %>%
  ungroup() %>%
  mutate(
    name = ifelse(name == "first", "1st", name),
    y = dense_rank(desc(earliest_day))
  )

ggplot(
  data = df_days_cerf,
  mapping = aes(
    x = days_from_allocation,
    y = y
  )
) +
  geom_rect(
    xmin = -Inf,
    xmax = 0,
    ymin = -Inf,
    ymax = Inf,
    fill = hdx_hex("sapphire-light")
  ) +
  geom_rect(
    xmin = 0,
    xmax = Inf,
    ymin = -Inf,
    ymax = Inf,
    fill = hdx_hex("tomato-light")
  ) +
  geom_text_hdx(
    aes(
      label = country
    ),
    color = "white",
    x = 0
  ) +
  geom_point(
    size = 8,
    color = hdx_hex("gray-dark")
  ) +
  geom_text_hdx(
    aes(
      label = name
    ),
    color = "white"
  ) +
  labs(
    y = "",
    x = "Days from CERF allocation",
    title = "Alerts relative to CERF allocation"
  ) +
  theme(
    axis.text.y = element_blank()
  ) +
  scale_x_continuous()
```
Overall, we can see that our alerts would have activated prior to nearly all CERF allocations. There are some outliers, such as in the DRC but looking at that country, it's unclear how and when outbreaks are calculated in the data CERF is scraping. Overall, we should feel pretty confident that we have some signal prior to activation.

```{r}
df_days_cerf %>%
  group_by(y, emergency_type, amount) %>%
  summarize(
    min_days = min(days_from_allocation),
    .groups = "drop"
  ) %>%
  ggplot(
    aes(
      x = min_days,
      y = amount,
      color = emergency_type
    )
  ) +
  geom_point() +
  scale_y_continuous_hdx(
    labels = scales::label_dollar()
  ) +
  expand_limits(y = 0) +
  scale_x_continuous() +
  coord_cartesian(
    clip = "off"
  ) +
  labs(
    y = "CERF allocation",
    x = "Time of first alert relative to CERF allocation",
    title = "Cholera alerts relative to CERF allocation size",
    color = "Emergency type"
  )
```
We can see that the alerts also generate earliest relative to the largest CERF allocations. This makes sense, as the longer CERF takes to respond, the more likely that any response would be to a much more widesrpread outbreak. Thus, the potential for speeding up allocations to new outbreaks by notifying CERF of their occurrence could have potentially high value.

```{r}
df_alert %>%
  group_by(iso3) %>%
  mutate(
    cases_max_26weeks = zoo::rollapply(
      data = cholera_cases,
      width = 26,
      FUN = max,
      na.rm = TRUE,
      fill = NA, 
      align = "left",
      partial = TRUE
    )
  ) %>%
  ungroup() %>%
  pivot_longer(
    alert_first:alert_10k,
    names_prefix = "alert_",
    names_to = "alert_type",
    values_to = "alert"
  ) %>%
  filter(
    alert
  ) %>%
  mutate(
    alert_type = factor(alert_type, levels = c("first", "1k", "5k", "10k"))
  ) %>%
  ggplot(
    aes(
      x = cases_max_26weeks
    )
  ) +
  geom_histogram(
    bins = 30
  ) +
  geom_vline(
    data = data.frame(
      cases_max_26_weeks = c(1000, 5000, 10000),
      alert_type = factor(c("1k", "5k", "10k"), levels = c("first", "1k", "5k", "10k"))
    ),
    aes(
      xintercept = cases_max_26_weeks
    ),
    color = "black"
  ) +
  facet_wrap(
    ~alert_type,
    scales = "free_y"
  ) +
  scale_x_continuous(
    labels = scales::label_comma()
  ) +
  labs(
    y = "",
    x = "# of cases",
    title = "Maximum # of cases in the 6 months following an alert",
    subtitle = "Initial alert threshold in black"
  )
```

From the above, it's clear that once an alert threshold reaches 5k or 10k, the
cholera cases generally tend to grow exponentially larger. This is less the case
for when there is the first cholera case and slightly less so for 1k threshold,
but still a relatively high percent. Let's have a look at those actual trends
from the date of alert to 6 months later

```{r, warning = FALSE}
df_alert %>%
  pivot_longer(
    alert_first:alert_10k,
    names_prefix = "alert_",
    names_to = "alert_type",
    values_to = "alert"
  ) %>%
  group_by(
    iso3, alert_type
  ) %>%
  mutate(
    alert_6_months = zoo::rollapply(
      data = alert,
      width = 26,
      FUN = any,
      na.rm = TRUE,
      align = "right",
      fill = NA,
      partial = TRUE
    ),
    alert_group = paste(iso3, cumsum(!alert_6_months))
  ) %>%
  filter(
    alert_6_months
  ) %>%
  group_by(
    alert_group
  ) %>%
  mutate(
    day = row_number() - 1,
    alert_type = factor(alert_type, levels = c("first", "1k", "5k", "10k"))
  ) %>%
  ggplot(
    aes(
      x = day,
      y = cholera_cases,
      group = alert_group
    )
  ) +
  geom_line(
    alpha = 0.2
  ) +
  facet_wrap(
    ~alert_type
  ) +
  scale_y_continuous_hdx(
    limits = c(0, 30000),
    labels = scales::label_comma()
  ) +
  labs(
    y = "Cholera cases",
    x = "Days since alert",
    title = "Cholera case trends following alert"
  )
  
```

It can still be hard to see the trends above, so let's simply look at the percent of alerts that are followed by another alert. That is, what % of alerts for the 1st case of cholera eventually see a future alert for 1,000 cases and so on.

```{r, fig.height = 2}
df_alert %>%
    group_by(iso3) %>%
    mutate(
        across(
            .cols = alert_1k:alert_10k,
            .fns = \(x) {
                zoo::rollapplyr(
                    data = x,
                    width = 26,
                    FUN = any,
                    na.rm = TRUE,
                    align = "left",
                    fill = NA,
                    partial = TRUE
                )
            },
            .names = "{.col}_6_months"
        )
    ) %>%
  pivot_longer(
    alert_first:alert_10k,
    names_prefix = "alert_",
    names_to = "alert_type",
    values_to = "alert"
  ) %>%
  mutate(
    alert_next_6_months = case_when(
      alert_type == "first" ~ alert_1k_6_months,
      alert_type == "1k" ~ alert_5k_6_months,
      alert_type == "5k" ~ alert_10k_6_months
    ),
    alert_type = factor(alert_type, levels = c("first", "1k", "5k"))
  ) %>%
  filter(
    alert,
    !is.na(alert_next_6_months)
  ) %>%
  group_by(
    alert_type
  ) %>%
  summarize(
    percent_6_months = sum(alert_next_6_months) / n()
  ) %>%
  ggplot(
    aes(
      y = alert_type,
      x = percent_6_months
    )
  ) +
  geom_bar(
    stat = "identity"
  ) +
  scale_x_continuous(
    labels = scales::label_percent()
  ) +
  labs(
    x = "% of alerts that escalated to the next alert",
    y = "",
    title = "Alerts that escalated in 6 months to the next alert",
    subtitle = "e.g. first case alert followed by 1k alert after 3 months"
  )
```

Above, we can see that 40% of alerts from a first case escalate in the next 6 months eventually to over 1,000 cases. Of course, we don't want to interpret these directly as true or false positives, but that 40% (and the increasing percents at higher levels) are quite high and indicate the value of alerting across these levels to ensure CERF is aware of any onsets or escalations.
